Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-01-21 ms.Rnw
more figures

\documentclass[10pt,reqno,final]{amsart}
\input{header}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,decorations,calc,math}

@@

<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
prefix <- "sgp"
source("setup.R")
@
<<packages,include=FALSE,cache=FALSE>>=
library(tidyverse)
library(ggtree)
library(pomp)
library(cowplot)
library(phylopomp)
stopifnot(getRversion() >= "4.3")
stopifnot(packageVersion("pomp")>="5.1")
stopifnot(packageVersion("phylopomp")>="0.9.2")
theme_set(theme_bw(base_family="serif"))
set.seed(1159254136)
@

@@

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \colorlet{darkblue}{blue!50!black}
        \tikzstyle{box}=[draw=black, text=black, fill=white, very thick, minimum size=3em]
        \tikzstyle{ibox}=[draw=darkblue]
        \tikzstyle{deme}=[fill=black!20!white]
        \tikzstyle{label}=[font=\Large]
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{flow}=[draw=black, very thick, >=stealth]
        \tikzstyle{modulate}=[draw=darkblue, >=Circle]

        %% SEIRS model
        \coordinate (origin) at (0,0);
        \node[label] (lab) at (origin) {\textbf{A}};
        \node [box] (S) at ($(origin)+(1,-1)$) {$\lab{S}$};
        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
        \node [box,ibox,deme] (I) at ($(E)+(2,0)$) {$\lab{I}$};
        \node [box] (R) at ($(I)+(2,0)$) {$\lab{R}$};
        \coordinate (overR) at ($(R)+(0,1)$);
        \coordinate (midSE) at ($(S)!0.5!(E)$);
        \draw [flow,->] (S) -- (E);
        \draw [flow,->] (E) -- (I);
        \draw [flow,->] (I) -- (R);
        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
        \draw [modulate,->] (I.north west) .. controls ($(I)+(-1,1)$) and ($(midSE)+(0,1)$) .. (midSE);
        \node[font=\normalsize] (demes) at ($(E)!0.5!(I)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I}\}$};

        %% SIIR (two-strain) model
        \coordinate (origin) at (0,-3);
        \node[label] (lab) at (origin) {\textbf{B}};
        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
        \node [box] (R1) at ($(S)+(6,1)$) {$\lab{R}_1$};
        \node [box] (R2) at ($(S)+(6,-1)$) {$\lab{R}_2$};
        \coordinate (overR) at ($(R1)+(0,1)$);
        \coordinate (underR) at ($(R2)-(0,1)$);
        \node [box,deme] (E1) at ($(S)+(2,1)$) {$\lab{E}_1$};
        \node [box,deme] (E2) at ($(S)+(2,-1)$) {$\lab{E}_2$};
        \node [box,ibox,deme] (I1) at ($(E1)+(2,0)$) {$\lab{I}_1$};
        \node [box,ibox,deme] (I2) at ($(E2)+(2,0)$) {$\lab{I}_2$};
        \draw [flow,->] (S) -- (E1);
        \draw [flow,->] (E1) -- (I1);
        \draw [flow,->] (I1) -- (R1);
        \draw [flow,->] (S) -- (E2);
        \draw [flow,->] (E2) -- (I2);
        \draw [flow,->] (I2) -- (R2);
        \draw [flow,->] (R1) -- (overR) -- (S |- overR) -- (S);
        \draw [flow,->] (R2) -- (underR) -- (S |- underR) -- (S);
        \coordinate (midSE1) at ($(S)!0.5!(E1)$);
        \coordinate (midSE2) at ($(S)!0.5!(E2)$);
        \draw [modulate,->] (I1.north west) .. controls ($(I1)+(-1,1)$) and ($(midSE1)+(0,1.8)$) .. (midSE1);
        \draw [modulate,->] (I2.south west) .. controls ($(I2)+(-1,-1)$) and ($(midSE2)+(0,-1.8)$) .. (midSE2);
        \node[font=\normalsize] (demes) at ($(E2)!0.5!(I2)+(0,-2)$) {$\Demes=\{\lab{E}_1,\lab{I}_1,\lab{E}_2,\lab{I}_2\}$};

        %% COVID model
        \coordinate (origin) at (9,0);
        \node[label] (lab) at (origin) {\textbf{C}};
        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
        \node [box] (R) at ($(S)+(7,1)$) {$\lab{R}$};
        \coordinate (overR) at ($(R)+(0,1)$);
        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
        \node [box,ibox,deme] (Ia) at ($(E)+(2,1)$) {$\lab{I_A}$};
        \node [box,ibox,deme] (Is) at ($(E)+(2,-1)$) {$\lab{I_S}$};
        \node [box] (H) at ($(Is)+(2,0)$) {$\lab{H}$};
        \node [box] (D) at ($(H)+(2,0)$) {$\lab{D}$};
        \draw [flow,->] (S) -- (E);
        \draw [flow,->] (E) -- (Ia.west);
        \draw [flow,->] (E) -- (Is.west);
        \draw [flow,->] (Ia) -- (R);
        \draw [flow,->] (Is) -- (H);
        \draw [flow,->] (H) -- (R);
        \draw [flow,->] (Is) -- (R);
        \draw [flow,->] (H) -- (D);
        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
        \coordinate (midSE) at ($(S)!0.5!(E)$);
        \draw [modulate,->] (Ia.north west) .. controls ($(Ia.north west)+(165:1)$) and ($(midSE)+(0,2)$) .. (midSE);
        \draw [modulate,->] (Is.south west) .. controls ($(Is.south west)+(195:1)$) and ($(midSE)+(0,-2)$) .. (midSE);
        \node[font=\normalsize] (demes) at ($(Is)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I_A},\lab{I_S}\}$};

        %% SI2R (super-spreader) model
        \coordinate (origin) at (9,-5);
        \node[label] (lab) at (origin) {\textbf{D}};
        \node [box] (S) at ($(origin)+(2,-1)$) {$\lab{S}$};
        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
        \node [box,ibox,deme] (Il) at ($(E)+(2,0)$) {$\lab{I_L}$};
        \node [box,ibox,deme] (Ih) at ($(Il)+(0,-2)$) {$\lab{I_H}$};
        \node [box] (R) at ($(Il)+(2,0)$) {$\lab{R}$};
        \coordinate (overR) at ($(R)+(0,1)$);
        \coordinate (midSE) at ($(S)!0.5!(E)$);
        \draw [flow,->] (S) -- (E);
        \draw [flow,->] (E) -- (Il);
        \draw [flow,<->] (Il) -- (Ih);
        \draw [flow,->] (Il) -- (R);
        \draw [flow,->] (Ih.east) -- (R |- Ih.east) -- (R);
        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
        \draw [modulate,->] (Il.north west) .. controls ($(Il.north west)+(135:0.8)$) and ($(midSE)+(0,1)$) .. (midSE);
        \draw [modulate,->] (Ih.west) .. controls ($(Ih.west)+(180:0.8)$) and ($(midSE)+(0,-1)$) .. (midSE);
        \node[font=\normalsize] (demes) at (midSE |- Ih) {$\Demes=\{\lab{E},\lab{I_L},\lab{I_H}\}$};
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Examples of discretely-structured population models.
    Demes are shaded.
    \textbf{(A)} An SEIRS model.
    Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
    Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
    If this immunity wanes, they re-enter the susceptible compartment.
    Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
    Accordingly, there are two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
    If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
    \label{fig:examples}
  }
\end{figure}

@@

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \usetikzlibrary{shapes,arrows,positioning}
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum height=8em]
        \tikzstyle{trans}=[color=black, font=\Large, thick, >=stealth]
        \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
        \node[state] (trans) at (165:8) {$x'=(S-1,E+1,I,R)$};
        \node[state] (prog) at (205:7) {$x'=(S,E-1,I+1,R+1)$};
        \node[state] (recov) at (305:5) {$x'=(S,E,I-1,R+1)$};
        \node[state] (wane) at (350:8) {$x'=(S+1,E,I,R-1)$};
        \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
        \draw[trans,->] (base) -- (trans) node[midway,above,sloped] {$\lab{Trans}$};
        \draw[trans,->] (base) -- (prog) node[midway,above,sloped] {$\lab{Prog}$};
        \draw[trans,->] (base) -- (recov) node[midway,above,sloped] {$\lab{Recov}$};
        \draw[trans,->] (base) -- (wane) node[midway,above,sloped] {$\lab{Wane}$};
        \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
        \node[font=\LARGE] (U) at (95:3.5) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Markov state transition diagram for the SEIRS model depicted in \cref{fig:examples}A.
    The state, $x$, is characterized by four numbers, $S$, $E$, $I$, and $R$.
    From a given state $x$, there are five possible kinds of jumps $x\mapsto{x'}$.
    Accordingly, the set, $\Jumps$, of jump marks has five elements.
    Each of these is of a different type:
    $\lab{Trans}$ (transmission) is of birth type,
    $\lab{Prog}$ (progression) is of migration type,
    $\lab{Recov}$ (recovery) is of death type,
    $\lab{Sample}$ (sampling) is of sample type,
    and $\lab{Wane}$ (loss or waning of immunity) is of neutral type.
    See \cref{sec:jump_types} for a description of these jump types.
    Note that, in this formulation, when a sampling event occurs, the state does not change.
    \label{fig:markov_state}
  }
\end{figure}

@@

\subsection{Genealogies}

For our purposes, a \emph{genealogy}, $G_{T}$, is a labeled, time-calibrated tree.
It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
There are three distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled living lineages;
\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
\item \emph{sample nodes}, which represent labeled samples.
\end{inparaenum}
Each node $a$ is associated with a specific time, $t(a)$.
In particular, if $a$ is a tip node in $G_T$, then $t(a)=T$;
if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
We let $\event{G}{T}$ denote the set of all node-times of the genealogy $G_T$.
Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.

Formally, we define a genealogy, $G_T$, to be an ordered pair $(y,z)$.
The $z$ element specifies the topological structure, as follows.
Let $\leaves{G_T}$ be the set of all tip and sample nodes in $G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{G_T}$.
The partition \emph{fineness} defines a partial order on $\func{P}$.
Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
The topological structure of $G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.

The $y$ element of $G_T$ specifies the coloring of branches and location of nodes.
In particular, for $a\in\leaves{G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
Thus $d:[0,T]\times\leaves{G_T}\to\Demesplus$ is a well-defined \cadlag\ function.

Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
Let $m:[0,T]\times\leaves{G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
Clearly, $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.

Finally, we define the map $y:[0,T]\times\leaves{G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
The two \cadlag\ functions $y$ and $z$ specify the genealogy $G_T$ completely and uniquely.
When the genealogy is not clear from context, we write $y^{G}$ and $z^{G}$ to refer to the respective components of genealogy $G$.

Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
In particular, $y$ must respect the topological structure of $G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
In other words, if $a,b\in{c}\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
Let $\Yspace^z$ denote the space of all functions that respect the topological structure described by $z$.
Let $\Yspace^\z_t$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.

@@

\begin{figure}
  \begin{center}
    \resizebox{0.8\linewidth}{!}{
      \begin{tikzpicture}[scale=2]
        \usetikzlibrary{shapes,arrows,positioning}
        \usetikzlibrary{arrows.meta,positioning,decorations}
        \colorlet{darkblue}{blue!50!black}
        \colorlet{maize}{yellow!90!black}
        \colorlet{grey}{black!30!white}
        \colorlet{darkgreen}{green!40!black}
        \colorlet{lightblue}{white!30!blue}
        \tikzstyle{label}=[font=\Large]
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{deme1}=[color=grey, line width=3pt, >={Circle[width=5pt,length=5pt]}]
        \tikzstyle{deme2}=[color=darkblue, line width=3pt, >={Circle[width=5pt,length=5pt]}]
        \tikzstyle{deme3}=[color=maize, line width=3pt, >={Circle[width=5pt,length=5pt]}]
        \tikzstyle{axis}=[color=black, thick, >=stealth]
        \tikzstyle{timepoint}=[color=red!70!black]

        \def\ysp{0.2} %% spacing between adjacent lineages

        %% Birth-type event
        \begin{scope}[shift={(0,0)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{A}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
          \coordinate (temp1) at ($(temp)+(0,\ysp)$);
          \coordinate (temp2) at ($(temp)-(0,\ysp)$);
          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
          \draw[deme3] (mid |- temp) -- (mid |- temp2) -- (right |- temp2);
          \draw[deme1] (left |- temp) -- (right |- temp);
          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
          \foreach \y in {1,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {6,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {2,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

        %% Death-type event
        \begin{scope}[shift={(2,0)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{B}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \foreach \y in {1,2,6} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {3,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
          \draw[deme3,dashed] (left |- temp) -- (mid |- temp);
          \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
          \foreach \y in {5,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

        %% Migration-type event
        \begin{scope}[shift={(4,0)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{C}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \foreach \y in {2,3,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
          \draw[deme2] (left |- temp) -- (mid |- temp);
          \draw[deme3] (mid |- temp) -- (right |- temp);
          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
          \foreach \y in {1,5} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {6,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

        %% Sample event
        \begin{scope}[shift={(6,0)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{D}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \foreach \y in {2,5} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {1,4,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {3,6,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \coordinate (temp) at ($(origin)+(0,-2*\ysp)$);
          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
          \coordinate (temp) at ($(origin)+(0,-6*\ysp)$);
          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

        %% Neutral-type event
        \begin{scope}[shift={(0,-2)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{E}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \foreach \y in {4,6,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {1,2,5} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {3,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

        %% Compound birth/death (Moran) event
        \begin{scope}[shift={(2,-2)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{F}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
          \coordinate (temp1) at ($(temp)+(0,\ysp)$);
          \draw[deme1] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
          \draw[deme1] (left |- temp) -- (right |- temp);
          \draw[deme3,dashed] (left |- temp1) -- (mid |- temp1);
          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
          \node[text=red,font=\Huge] at (mid |- temp1) {$\times$};
          \foreach \y in {2,5} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {1,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {6,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

        %% Compound sample/death event
        \begin{scope}[shift={(4,-2)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{G}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \coordinate (temp) at ($(origin)+(0,-5*\ysp)$);
          \draw[deme3] (left |- temp) -- (mid |- temp);
          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
          \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
          \foreach \y in {1,6} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {2,3,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {4,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

>>>>

\documentclass[10pt,reqno,final]{amsart}

\input{header}

@@

<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
prefix <- "sgp"
source("setup.R")
@
<<packages,include=FALSE,cache=FALSE>>=
library(tidyverse)
library(ggtree)
library(pomp)
library(cowplot)
library(viridis)
library(phylopomp)
stopifnot(getRversion() >= "4.3")
stopifnot(packageVersion("pomp")>="5.1")
stopifnot(packageVersion("phylopomp")>="0.10.1")
theme_set(theme_bw(base_family="serif"))
set.seed(1159254136)
@

@@

\begin{figure}
  \input{figs/example_models}
\end{figure}

@@

\begin{figure}
  \input{figs/markov_diagram}
\end{figure}

@@

\subsection{Genealogies}
\label{sec:genealogy}

For our purposes, a \emph{genealogy}, $\G_{T}$, is a labeled, time-calibrated tree.
It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
There are three distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled living lineages;
\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
\item \emph{sample nodes}, which represent labeled samples.
\end{inparaenum}
Each node $a$ is associated with a specific time, $t(a)$.
In particular, if $a$ is a tip node in $\G_T$, then $t(a)=T$;
if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
We let $\event{\G}{T}$ denote the set of all node-times of the genealogy $\G_T$.
Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.

Formally, we define a genealogy, $\G_T$, to be an ordered pair $(y,z)$.
The $z$ element specifies the topological structure, as follows.
Let $\leaves{\G_T}$ be the set of all tip and sample nodes in $\G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{\G_T}$.
The partition \emph{fineness} defines a partial order on $\func{P}$.
Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
The topological structure of $\G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.

The $y$ element of $\G_T$ specifies the coloring of branches and location of nodes.
In particular, for $a\in\leaves{\G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
Thus $d:[0,T]\times\leaves{\G_T}\to\Demesplus$ is a well-defined \cadlag\ function.

Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
Let $m:[0,T]\times\leaves{\G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
Clearly, $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.

Finally, we define the map $y:[0,T]\times\leaves{\G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
The two \cadlag\ functions $y$ and $z$ specify the genealogy $\G_T$ completely and uniquely.
When the genealogy is not clear from context, we write $y^{\G}$ and $z^{\G}$ to refer to the respective components of genealogy $\G$.

Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
In particular, $y$ must respect the topological structure of $\G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
In other words, if $a,b\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
Let $\Yspace^z$ denote the space of all functions that obey this constraint at every $t$.
Let $\Yspace^\z_t\subseteq{\Demesplus\times\Zp}$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.

@@
